---
title: "Introduction to R: Part 2-A"
author: "Jenna Krall, PhD"
date: "January 2016"
height: 900
font-family: Helvetica
width: 1440
fontsize: 10pt
output:
  beamer_presentation:
    includes:
      in_header: images/header.tex
---



```{r set_knitr_chunk_options, echo = F, message = F, warning = F}
# Source file with setup
source("sourcefile.R")
```




Modules Part 2
========================================================

1. Merging and transforming data
2. ggplot2
3. Control structures and functions
4. Reproducibility in R 
5. R extensions  


Module 2-1
========================================================

Merging/transforming data

1. Merging datasets
2. Transforming datasets



Merging and transforming datasets
========================================================

```{r mod1}
vlbw <- read.csv("vlbw.csv", stringsAsFactors = F)
vlbw <- select(vlbw, bwt, gest, ivh, delivery, pneumo, twn) 
vlbw <- vlbw[complete.cases(vlbw), ]
vlbw <- mutate(vlbw, id = 1 : nrow(vlbw))
```



Merging and transforming datasets
========================================================

- Suppose we had some additional data on birthweight at week 1 and week 2 after birth
- Infants gain approximately 150 grams per week
```{r}
set.seed(1807)
n <- nrow(vlbw)
weight1 <- vlbw$bwt + 150 + rnorm(n, sd = 10) 
weight2 <- weight1 + 150 + rnorm(n, sd = 10)
weights <- data.frame(seq(1, nrow(vlbw)), weight1, weight2)
colnames(weights) <- c("id", "weight_1", "weight_2")
weights <- tbl_df(weights)
```



Merging and transforming datasets
========================================================

```{r}
weights
```



Merging and transforming datasets
========================================================
We can merge these two datasets using the `join` functions

```
full_join(x = data1, y = data2, by = what variable to merge by?)
```

```{r}
m_vlbw <- full_join(vlbw, weights, by = "id")
head(m_vlbw)
```






Merging and transforming datasets
========================================================
If column names of ID are not the same, can specify `by` argument:

```{r}
weights2 <- rename(weights, infantID = id)
head(weights2)
```


Merging and transforming datasets
========================================================
If column names of ID are not the same, can specify `by` argument:

- `id` is name of ID in `vlbw`
- `infantID` is name of ID in `weights`

```{r, eval = T}
m_vlbw <- full_join(vlbw, weights2, by = c("id" = "infantID"))
head(m_vlbw)
```




Merging and transforming datasets
========================================================
- IDs may differ between the datasets

```{r}
# Restrict to IDs less than 35
vlbw_2 <- filter(vlbw, id < 35)
n_distinct(vlbw_2$id)
n_distinct(vlbw$id)
```

Merging and transforming datasets
========================================================
- IDs may differ between the datasets
- Left join only keeps IDs in first dataset 
```{r}
m_vlbw <- left_join(vlbw_2, weights)
n_distinct(m_vlbw$id)
```



**No data for IDs 35-671**








Exercise
========================================================

Exercise 1a

- Use the function `left_join` to merge `vlbw_2` and `weights` to create a dataset with only IDs that are in the weights dataset








Transforming datasets
========================================================


These data are in what we call "wide form".  

- Each ID corresponds to one row in the dataset
- Columns represent the same variable over time (weights)


Transforming datasets to "long form"

- Each row represents a unique observation.  
- Package `tidyr`

```
gather(data, key = timevar, value = valuevar, what switches)
```


Transforming datasets
========================================================


```{r}
head(m_vlbw)
long_data <- gather(m_vlbw, key = time, value = weight, weight_1 : weight_2)
head(long_data)
```



Transforming datasets
========================================================

- Suppose we also have height data over time
- Infants gain approximately 1/4 inch per week
```{r}
set.seed(831)
n <- nrow(vlbw)
height1 <- 5 + 0.25 + rnorm(n, sd = 0.01) 
height2 <- height1 + 0.25 + rnorm(n, sd = 0.01)
heights <- data.frame(seq(1, nrow(vlbw)), height1, height2)
colnames(heights) <- c("id", "height_1", "height_2")
heights <- tbl_df(heights)
```

Transforming datasets
========================================================

- Infants gain approximately 1/4 inch per week
```{r}
heights
```



Transforming datasets
========================================================

Transform heights and weights into long form
```{r}
hw <- full_join(heights, weights)
head(hw)
```

Transforming datasets
========================================================

```{r}
long_data2 <- gather(hw, key = key, value = value, height_1 : weight_2)
head(long_data2)

```


Transforming datasets
========================================================

- This is now a `tidy` dataframe
```{r}
tidy_data <- separate(long_data2, key, into = c("variable", "time"), 
    sep = "_")
head(tidy_data)
```


Transforming datasets
========================================================

```{r}
table(long_data2$key)
table(tidy_data$variable)
table(tidy_data$time)
```











Transforming datasets
========================================================
Returning the data to "long form",
```
spread(data, key = timevar, value = valuevar)
```


```{r}
wide_data <- spread(long_data, key = time, value = weight)
head(wide_data)
```



tidyr Rstudio cheatsheet
========================================================

\includegraphics[scale=.35]{images/tidyr.png}













Module 2-2
========================================================

Package `ggplot2`

1. Recreating simple R plots in ggplot2
2. Facetting




Basic ggplots
========================================================

Scatterplot

```{r mod2}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) + geom_point()
```



Basic ggplots
========================================================

Color all points
```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) + 
    geom_point(color = "red")
```



Basic ggplots
========================================================

Change shape of all points
```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length)) + 
    geom_point(shape = 2)
```



Basic ggplots
========================================================

Color by factor variable

- Automatic legend
```{r, echo = F}
opts_chunk$set(out.width= '8cm', fig.width = 6)
```

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) + 
    geom_point()
```






Basic ggplots
========================================================


Shape by factor variable
- Automatic legend
```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point()
```






Basic ggplots
========================================================

Titles and labels

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() + 
    xlab("Sepal width") + ylab("Sepal length") +
    ggtitle("Fisher's iris data")
```


Facetting
========================================================

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() +
    facet_wrap(~ Species)
```




Facetting
========================================================

To change facet labels, or change facet order, use `factor`

```{r, eval = F}
iris2 <- iris
iris2$Species <- factor(iris$Species, levels = c("versicolor",
    "virginica", "setosa"), labels = c("VER", "VIR", "SET"))

ggplot(iris2, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() +
    facet_wrap(~ Species)
```


Facetting
========================================================

To change facet labels, or change facet order, use `factor`

```{r, echo = F}
iris2 <- iris
iris2$Species <- factor(iris$Species, levels = c("versicolor",
    "virginica", "setosa"), labels = c("VER", "VIR", "SET"))

ggplot(iris2, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() +
    facet_wrap(~ Species)
```






Facetting
========================================================

```{r}
iris <- mutate(iris, petlencut = cut(Petal.Length, breaks = 4))
table(iris$petlencut)
```




Facetting
========================================================

```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() +
    facet_wrap(petlencut~ Species)

```



Exercise
========================================================

Exercise 2a

- Plot sepal length against sepal width, and color the points by petlencut


Facetting
========================================================

Facet grid
```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species,
    shape = Species)) + geom_point() +
    facet_grid(petlencut~ Species)

```




Colors
========================================================

Coloring by a discrete variable
```{r}
cols <- c("black", "red", "green")
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Species)) +   
    geom_point() +
    scale_color_manual(values = cols)
```





Colors
========================================================

Coloring by a continuous variable
```{r}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, color = Petal.Length)) +   
    geom_point() +
    scale_color_gradient(low = "black", high = "red")
```





Line plots
========================================================

```{r}
data(airquality)
aq1 <- filter(airquality, Month == 5)
head(aq1)
```



Line plots
========================================================

```{r}
ggplot(aq1, aes(x = Day, y = Ozone)) + geom_line()
```


Line plots
========================================================

```{r}
ggplot(aq1, aes(x = Day, y = Ozone)) + geom_line(linetype = 2)
```




Line plots
========================================================

```{r}
aq1 <- mutate(airquality, Month = factor(Month))
ggplot(aq1, aes(x = Day, y = Ozone)) + 
    geom_line(aes(linetype = Month))
```



Exercise 2b
========================================================

- Change the previous plot of ozone by day so that the color of the lines varies by month

Histograms
========================================================


```{r}
ggplot(iris, aes(x = Sepal.Width)) + geom_histogram() 
```

Histograms
========================================================

```{r}
ggplot(iris, aes(x = Sepal.Width, fill = Species)) + geom_histogram() 

```








Barplots
========================================================


```{r}
ggplot(iris, aes(x = Species)) + geom_bar() 
```

Barplots
========================================================

```{r}
head(state.x77)
state <- data.frame(rownames(state.x77), state.x77)
colnames(state)[1] <- "State"
```


Barplots
========================================================

```{r, fig.width = 8, out.width = "10cm"}
ggplot(state, aes(x = State, y = Illiteracy, fill = State)) + 
    geom_bar(stat = "identity") 
```


Boxplots
========================================================
```{r}
ggplot(iris, aes(Species, Sepal.Length)) + geom_boxplot()
```





Plotting confidence intervals
========================================================

Creating a dataset of regression results
```{r}
iris_spec <- group_by(iris, Species)  
means <- summarise_each(iris_spec, funs(mean), Sepal.Length : Petal.Width)
means <- gather(means, "Variable", "Mean", 2 : 5)
sds <- summarise_each(iris_spec, funs(sd), Sepal.Length : Petal.Width) 
sds <- gather(sds, "Variable", "SD", 2 : 5)
```



Exercise
========================================================

Exercise 2c

- Use `full_join` to join together means and standard deviations to create one datastet
- Use `mutate` to create two new variables representing mean - 1 SD and mean + 1 SD






```{r, echo = F}

CIdat <- full_join(means, sds)
CIdat <- mutate(CIdat, LB = Mean - SD, UB = Mean + SD)

```

 
Plotting intervals
========================================================
```{r}
CIdat
```

Plotting intervals
========================================================


Using geom_pointrange
```{r}
ggplot(CIdat, aes(y = Mean, x = Species, color = Species)) +
    geom_pointrange(aes(ymin = LB, ymax = UB)) +
    facet_wrap(~ Variable)
```



Plotting intervals
========================================================

Using geom_pointrange
```{r}
ggplot(CIdat, aes(y = Mean, x = Species, color = Species)) +
    geom_pointrange(aes(ymin = LB, ymax = UB)) +
    facet_wrap(~ Variable, scales = "free")
```



Plotting intervals
========================================================

Using geom_pointrange
```{r}
ggplot(CIdat, aes(y = Mean, x = Species, color = Species)) +
    geom_pointrange(aes(ymin = LB, ymax = UB)) +
    facet_wrap(~ Variable, scales = "free_y")
```



Plotting intervals
========================================================

Using geom_errorbar
```{r}
ggplot(CIdat, aes(y = Mean, x = Species, color = Species)) +
    geom_errorbar(aes(ymin = LB, ymax = UB)) +
    facet_wrap(~ Variable, scales = "free_y")
```




Exercise
========================================================


Exercise 2d

- Recreate the plot above using `geom_errorbar`, but removing the horizontal lines at each bar, and adding in the point estimates




Plotting intervals
========================================================

```{r}
ggplot(CIdat, aes(y = Mean, x = Species)) +
    geom_errorbar(aes(ymin = LB, ymax = UB,color = Species)) +
    geom_point(color = "red") +
    facet_wrap(~ Variable, scales = "free_y")
```





Themes
========================================================

- Themes control sizing, color, etc.
- Are "added" to ggplot
- Syntax:

```
theme(thing we want to change = theme attribute(attribute assigning))
```


Themes
========================================================

```{r}
g1 <- ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, 
    color = Species)) + geom_point()
g1
```



Themes
========================================================

- Theme attribute: `element_text`
- Changing the size of all titles

```{r}
g1 + theme(title = element_text(size = 25))
```






Themes
========================================================

- `element_line` is used to change lines in ggplot
```{r}
g1 + theme(axis.ticks.y = element_line(size = 2))
```



Themes
========================================================

- Changing the background
```{r}
g1 + theme_bw()
```


Themes
========================================================

- Changing the background
```{r}
g1 + theme_classic()
```


Themes
========================================================


- `element_blank` is used to remove items from the plot
```{r}
g1 + theme(axis.text = element_blank())
```





Legends
========================================================

- Removing legend
```{r}
g1 + theme(legend.position = "none")
```


Legends
========================================================

- Move legend to top
```{r}
g1 + theme(legend.position = "top")
```










RColorBrewer
========================================================

- Package `RColorBrewer`

```{r}
cols <- brewer.pal(n = 5, name = "Dark2")
g1 + scale_color_manual(values = cols)
```



RColorBrewer
========================================================

- `ggplot` can do this easily:

```{r}
g1 + scale_colour_brewer(palette = "Dark2")
```



Saving ggplots
========================================================


Same as base plotting, though can use saved objects
```{r, eval = F}
png("plot1.png")
g1
dev.off()
```

ggplot Rstudio cheatsheet
========================================================

\includegraphics[scale=.35]{images/tidyr.png}






























Module 2-3
========================================================


Control structures and functions


1. If/then statements
2. For loops
3. Apply statements
4. Functions






Control structures
========================================================

"Controlling" how R performs

- Usually, code in R is executed sequentially (line 1, line 2, etc.)
- Control structures are useful when we only want R to execute code if a certain condition is met, or if we want to do the same procedure multiple times


In SAS

- `if`/`then` statements
- `do` loops

In R

- `if`/`then` statements
- `for`, `while` loops





If/then/else
========================================================
Evaluate code based on whether a condition is true

- If the species is setosa, make the color blue

Syntax:
```
if ( conditional statement ) do this
```

Example 1:
```{r}
spec1 <- iris$Species[1]
spec1 == "setosa"
if (iris$Species[1] == "setosa") {
    print("The species is setosa")
    }
```





If/then/else
========================================================

- Setting conditions when if statement is not true


- if/then/else
```{r}
j <- 51
if (iris$Species[j] == "setosa") {
    print("The species is setosa") 
}else {
    "Not setosa"
}         
```

If/then/else
========================================================



- `ifelse` 
- Syntax: ifelse(condition, true result, false result)
```{r}
if1 <- ifelse(iris$Species[j] == "setosa", "The species is setosa", "Not setosa")

if1
```





If/then/else
========================================================

Can create multiple `if/else`'s

Syntax:
```
if (conditional statement) {
    do thing 1
    do thing 2
} else if (second conditional statement) {
    do thing 3 
    do thing 4
} else if () {
    do...
} else if () {
    do...
} else {
    catchall...
    last thing!
}
```


If/then/else
========================================================


```{r}
if (iris$Species[j] == "setosa") {
    print("The species is setosa")
} else if (iris$Species[j] == "virginica") {
    print("The species is virginica")
} else {
    print("The species is versicolor")
}
```



If/then/else
========================================================

- `else` can lead to potential errors
```{r}
iris$Species[101]
j <- 101
if (iris$Species[j] == "setosa") {
    print("The species is setosa")
} else if (iris$Species[j] == "Virginica") {
    print("The species is virginica")
} else {
    print("The species is versicolor")
}
```




If/then/else
========================================================

Be careful with `else`!
```{r}
iris$Species[101]
j <- 101
if (iris$Species[j] == "setosa") {
    print("The species is setosa")
} else if (iris$Species[j] == "Virginica") {
    print("The species is virginica") 
} else if(iris$Species[j] == "versicolor") {
    print("The species is versicolor")    
} else {
    print("Error: species not defined")
}
```




Exercise
========================================================

Exercise 3a

- Reread the "vlbw.csv" dataset, calling it vlbw2
- Create the outcome variable using mutate, being careful to categorize missing IVH as missing outcome (Hint: use two `ifelse()` commands)







For loops
========================================================

Performing the same task for multiple variables or conditions

- Take the mean of age, years of education, income, other variables
- Make scatterplots of CD4 count against a series of risk factors

Syntax:
```
for (how will we loop?) do this
for (how will we loop?) {
    do these things
}
```




For loops
========================================================

"How will we loop?"

- We need an index (frequently `i`)
- We need to say how that index will change over the loop
- e.g. `i in 1 : 4`

For loops
========================================================

Example 1:
```{r}
for (i in 1 : 4) {
    print(i)
}
for (i in c(1, 2, 5)) {
    print(i)
}
```


For loops
========================================================

Example 2:
```{r}
seq(1, 10, by = 2)
for (i in seq(1, 10, by = 2)) {
    print(i)
}

```


For loops
========================================================
Example 3:

```{r, echo = F}
set.seed(10)
matrix_1 <- matrix(round(rnorm(20, mean = 5, sd = 2), 2), nrow = 4)
```
```{r}
matrix_1
for (i in c(1, 3)) {
    print(matrix_1[i, ])
}

```




For loops
========================================================

Example 4: column means
```{r}
head(iris)
for (i in 1 : 4) {
    print(head(iris[, i]))
    mean(iris[, i])
}

```






For loops
========================================================

- Saving our means
```{r}
mean_iris <- vector(length = 4)
mean_iris
for (i in 1 : 4) {
    mean_iris[i] <- mean(iris[, i])
}
mean_iris

```







For loops
========================================================


Useful if have lots of steps and don't need to save intermediate steps

```{r}
vlbw <- mutate(vlbw, outcome = 1 * ((ivh %in% c("definite", "possible"))))
for(i in colnames(vlbw)) {
    print(i)
    if(!(i %in% c("outcome", "ivh", "id"))) {
        eqn1 <- paste("outcome ~ ", i)
        print(eqn1)
        glm1 <- glm(eqn1, data = vlbw, family = "binomial")
    }
}
```




For loops
========================================================


```{r}
glmout <- data.frame(colnames(vlbw), matrix(ncol = 2, 
    nrow = length(colnames(vlbw))))
colnames(glmout) <- c("Variable", "Est", "SE")
k <- 1
for(i in colnames(vlbw)) {
    if(!(i %in% c("outcome", "ivh", "id"))) {
        eqn1 <- paste("outcome ~ ", i)
        glm1 <- glm(eqn1, data = vlbw, family = "binomial") 
        glmout[k, 2 : 3] <- summary(glm1)$coef[-1, c(1, 2)]
    }
    k <- k + 1
}
glmout <- mutate(glmout, LB = Est - 1.96 * SE, 
    UB = Est + 1.96 * SE)
```






For loops
========================================================


```{r}
glmout
```


Exercise 
========================================================

Exercise 3b

- Construct a for loop for creating scatterplots in base R of sepal length vs. the remaining numeric variables (sepal width, petal length, petal width), including correct axis labels. 
- Hint: might be easiest to use brackets to extract columns from the iris dataset



Apply
========================================================

Series of R functions that automate loops

- apply: apply a function to margins of matrix (rows or columns)
    - e.g. Means of variables (columns)
    - e.g. Standard deviation of observations (rows)
- tapply: apply a function separately to groups of an indexing variable
    - e.g. Mean heart rate by sex
    - e.g. Median CD4 count by income group
- lapply: apply a function to each element of a list
- sapply: same as lapply, but tries to simplify




Apply
========================================================
Syntax for apply:
```
apply( X = mymatrix, MARGIN = 1 or 2, FUN = function of choice)
```
- `MARGIN = 1` corresponds to rows
- `MARGIN = 2` corresponds to columns

```{r}
matrix_1 <- matrix(rnorm(40), nrow = 4)
dim(matrix_1)
```

Apply
========================================================
Finding standard deviation of each row:
```{r}
sd_out <- vector(length = 4)
sd_out[1] <- sd(matrix_1[1, ])
sd_out[2] <- sd(matrix_1[2, ])
sd_out[3] <- sd(matrix_1[3, ])
sd_out[4] <- sd(matrix_1[4, ])
sd_out
```

Apply
========================================================

```{r}
sd_out <- vector(length = 4)
for(i in 1: 4) {
    sd_out[i] <- sd(matrix_1[i, ])
}
sd_out
apply(matrix_1, 1, sd)
```

Apply
========================================================
Finding median of each column:
```{r}
apply(matrix_1, 2, median)
```


Apply
========================================================


We can explicitly use for loops for finding column means:
```{r}
mean_iris <- vector()
for (i in c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width")) {
    mean_iris[i] <- mean(iris[, i])
}
mean_iris
```

Apply
========================================================


One line of code, same result
```{r}
apply(iris[, 1 : 4], 2, mean)
summarise_each(iris, funs(mean), Sepal.Length : Petal.Width)
```
*Summarize more useful when have a grouping variable*

Apply
========================================================

Can pass other arguments to mean
```{r}
apply(iris[, 1 : 4], 2, mean)
apply(iris[, 1 : 4], 2, mean, trim = 0.1, na.rm = T)
```




Apply
========================================================

`tapply` allows us to apply a function by a factor

- e.g. What is the mean sepal length by species?
- e.g. What is the mean FEV by age category?

Syntax:
```
tapply(thing to take FUN of, factor, FUN)
```


Apply
========================================================

Summarizing irises by species
```{r}
tapply(iris$Sepal.Length, iris$Species, median)
tapply(iris$Sepal.Width, iris$Species, sd)
tapply(iris$Sepal.Width, iris$Species, length)
```



Exercise
========================================================

Exercise 3c

- Compute the mean sepal length by species using

    1. a for loop
    2. `tapply`
    3. `summarise` using dplyr












Functions
========================================================

Syntax

```
myfun <- function(argument1, argument2, argumentN) {
    do things
}
```



Functions
========================================================

Function for the quadratic equation
```{r mod4}
quadeq <- function(a, b, c) {
    num1 <- (-b + sqrt(b^2 - 4 * a * c)) 
    num2 <- (-b - sqrt(b^2 - 4 * a * c)) 
    den <- 2 * a
    output <- cbind(num1 / den, num2 / den)
    return(output)
}
quadeq(1, 3, 2)
quadeq(1, 4, 2)

```




Functions
========================================================

Function arguments
```{r}
newsummary <- function(x) {
    out <- c(mean(x), sd(x), min(x), max(x))
    return(out)
}
newsummary(rnorm(10))
```




Functions
========================================================

Function for nicer base R plot

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width, col = iris$Species)
```

Functions
========================================================

Function to set better plot defaults

```{r}
nicescatter <- function(data, xvar, yvar, colorvar) {
    plot(data[, xvar], data[, yvar], col = data[, colorvar], pch = 16)
}
```


Functions
========================================================

Function to set better plot defaults

```{r}
nicescatter(iris, "Sepal.Length", "Sepal.Width", "Species")
```



Functions
========================================================


Specifying defaults

```{r}
nicescatter <- function(data, xvar, yvar, cols = "red") {
    plot(data[, xvar], data[, yvar], col = cols, pch = 16)
}
```


Functions
========================================================


Specifying defaults

```{r}
nicescatter(iris, "Sepal.Length", "Sepal.Width")
```





Functions
========================================================


Specifying defaults

```{r}
nicescatter <- function(data = iris, xvar = "Sepal.Length", 
    yvar = "Sepal.Width", colorvar = "Species") {
    plot(data[, xvar], data[, yvar], col = data[, colorvar], pch = 16)
}
```




Functions
========================================================


Specifying defaults

```{r}
nicescatter()
```


Functions
========================================================


Specifying defaults

```{r}
nicescatter(xvar = "Petal.Width")
```



Exercise
========================================================

Exercise 3d

- Modify the `nicescatter` function to add x- and y-axis labels, and a legend automatically




Functions
========================================================


- R will try and match arguments by name, and then by order
- Safest approach is to always specify `argument = value` within function call




Functions
========================================================

Central limit theorem example

```{r}
# disttype distribution type
# n size of samples
# N number of samples
clt <- function(disttype, n, N) {
    if(disttype == "normal") {
        x <- rnorm(n * N)
    } else if(disttype == "poisson") {
        x <- rpois(n * N, lambda = 2)
    } else if(disttype == "binomial") {
        x <- rbinom(n * N, size = 1, prob = 0.5)
    } else {
        stop("Distribution not found")
    }
    x <- matrix(x, ncol = n)
    output <- apply(x, 1, mean)
}
```



Syntax
========================================================

```{r}
x <- clt("normal", 100, 100)
hist(x)
```


Syntax
========================================================

```{r}
x <- clt("poisson", 100, 100)
hist(x)
```


Syntax
========================================================

```{r}
x <- clt("binomial", 1000, 1000)
hist(x)
```







Overview exercise 3
========================================================



Using the airquality dataset,

1. Create a date variable in MM-DD-YY (Hint: the data are from 1973, the functions `unite`, `paste0`, and `as.Date` might be useful)




```{r}
# Create Date variable
airquality <- mutate(airquality, Month = paste0("0", Month), 
    Day = ifelse(Day < 10, paste0("0", Day), Day))
airquality <- unite_(airquality, "Date", c("Month", "Day"), sep = "-", remove = F)
airquality <- mutate(airquality, Date = as.Date(paste0("1973-", Date)))
```



Overview exercise 3
========================================================


2. Transform the data into `tidy` form (i.e. one column corresponding to variable, one column for the value of that variable (Hint: `gather`))
3. Plot the time series of each variable (e.g. Ozone vs. Date), faceting on variable (Hint: `geom_line`)
4. Compute the mean, minimum, and maximum value for each variable (Ozone : Temp) for each day, across months (Hint: `group_by` and `summarise`)
5. Plot the mean variable value against day, by faceting on each variable.  Additionally, add a ribbon (`geom_ribbon`) for the minimum and maximum value.





