---
title: "Introduction to R: Part 2-B"
author: "Jenna Krall, PhD"
date: "January 2016"
height: 900
font-family: Helvetica
width: 1440
fontsize: 10pt
output:
  beamer_presentation:
    includes:
      in_header: images/header.tex
---



```{r set_knitr_chunk_options, echo = F, message = F, warning = F}
# Source file with setup
source("sourcefile.R")
```



Modules Part 2
========================================================

1. Merging and transforming data
2. ggplot2
3. Control structures and functions
4. Reproducibility in R 
5. R extensions  




Module 2-4
========================================================

Reproducibility in R

1. Introduction to reproducible research
2. External code
3. `knitr`
4. Code chunk options
5. Tables, figures, in-line code



Reproducible research
========================================================

``...the calculation of quantitative scientific results by independent scientists using the original datasets and methods''

What do we need?

- Reproducibility = Tools + Workflow (Stodden, Leisch, and Peng)
- Literate statistical programming (Rossini via Knuth):
    - programming language + documentation language
- We need to link **data, code, results, and interpretation**








Sourcing external code
========================================================
We might have 

- One .R file to clean and merge datasets
- One .R file to create output including tables and figures

If we want to separate the two files, we can call the second from the first using the `source` function.
```{r mod5}
rm(list = ls())
ls()
```
```{r, eval = T}
source("data_clean.R")
```

```{r}
ls()
```

```{r, echo = F, eval = T}
echo1 <- F
```


Introduction to knitr
=======================================================

knitr

- Developed by Yihui Xie at Iowa State (now at RStudio)
- Integration with RStudio

We will focus on using knitr to integrate R and markdown, but knitr is actually a flexible tool that can be used in broader ways.

- We write text (in markdown) separated by code chunks (written in R)
- In our output, we include code (or not) and results (relevant tables and figures)
- knitr ties together our code and our final report





This class is reproducible
========================================================

**ALL** slides for this class have been created using knitr.  This means that if R changes, if the dataset changes, if you want to see how I created the slides, *the code and output and interpretation is all there*.



Ex 1: Plotting
========================================================

```{r, echo = echo1}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, 
    color = Species)) + geom_point()
```


Ex 1: Plotting
========================================================

```{r, echo = !echo1}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length, 
    color = Species)) + geom_point()
```



Ex 2: Summary statistics
========================================================

```{r, results="verbatim", echo = echo1}
summary(iris)
```



Example 2: Summary statistics
========================================================

```{r, results="verbatim", echo = !echo1}
summary(iris)
```





What is markdown?
========================================================

- Markdown is a markup language
- Use to write reports, papers, etc.
- Replaces "buttons" in Microsoft Word or similar with code

### How to format using markdown

- Preceeding a word with "#" or "##" makes it a header and subheader respectively
- Surrounding a word with "\*" or "\*\*" makes it *italics* or **bold** respectively
- Bullets are created by a dash and then a space
- Indented bullets are created by using two spaces before the dash
  
  
1. Numbered lists are created by specifying the number, a period, and a space
2. Here is number 2.





How to use knitr
========================================================

### Filling in the content


- Write some text using markdown, e.g. ``I am doing an analysis of the iris dataset in R.''
- Add code in ``code chunks''
    - Start with three backticks and a bracketed \{r\}
    - End with three backticks 


\```{r} 
```
mean(rnorm(10))
```
\```

When you are finished, click the knit button to create your dynamic document.




Code chunk options and names
========================================================


### Code chunk names
- Code chunks can be named within the bracket: {r codechunk1} or {r simulatedata}
- Useful to indicate what the code chunk does
- Can help with debugging documents
- Can make it easier to navigate a long document


  
  
  
Code chunk options and names
========================================================

You can specify code chunk options like arguments to an R function

- Code chunk options are specified after a comma following `r` in the code chunk.
- Useful code chunk options:
    - eval: TRUE or FALSE indicating whether you want R to run the code. 
    - echo: TRUE or FALSE indicating whether the final document should show the code
e.g. 

\```{r codechunk1, echo = T, eval = F}

`mean(rnorm(10))`

\```


  
Code chunk options and names
========================================================


- warning = TRUE or FALSE depending on whether you want to display warnings
- message = TRUE or FALSE depending on whether you want to display messages

```{r, warning = T, message = T}
detach(package:dplyr)
library(dplyr)
```
  
Figures in knitr
========================================================  
  
  
We can include figures in `knitr` using the same base R or ggplot functions

- You can control the output with code chunk options out.width and out.height, which specify the height and width of the chunk in pixels (for html file)

```{r dataclean, verbatim = T}
g1 <- ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) + 
    geom_point() 
```




  
Figures in knitr
========================================================  
 
```{r}
g1
```

Tables in knitr
========================================================  

```{r, echo = T}
iris_spec <- group_by(iris, Species)
s1 <- summarise_each(iris_spec, funs(mean), Sepal.Length : Petal.Width)
s1
```



Tables in knitr
========================================================  

Convenient and easy to use knitr to make tables by using the `kable` function

```{r, echo = T}
kable(s1)
```



Tables in knitr
========================================================  

We can also use the arguments of kable
```{r}
kable(s1, digits = 2)
```

Tables in knitr
========================================================  

We can also add captions and realign the output
```{r}
kable(s1, digits = 0, align = rep("c", 4), format = "pandoc",
  caption = "Mean of iris variables by species")
```




In-line code
========================================================

You can put code in the lines of your text by using:

\`{r mean(var1)}`

As an example, here is my code chunk that defines date:
```{r, verbatim = T}
date <- Sys.Date()
```

Now I can do inline code specifying the day: Today is `r date`.  Now, when I rerun the code on a different day, this will update.

Include code that you don't want to run inside two backticks, e.g. today we are learning about `knitr`.   



Ex 3: Writing interpretations.
========================================================
The R dataset *iris* has `r ncol(iris)` variables, including `r paste(colnames(iris), collapse = ", ")`.  The dataset has `r nrow(iris)` observations.  The average petal length is `r mean(iris$Petal.Length)` with a standard deviation (sd) of `r sd(iris$Petal.Length)`.  The average petal width was `r mean(iris$Petal.Width)` (sd = `r sd(iris$Petal.Width)`).  




Ex 3: Writing interpretations.
========================================================
```{r, echo = !echo1}
iris <- filter(iris, Species != "setosa")
```
The R dataset *iris* has `r ncol(iris)` variables, including `r paste(colnames(iris), collapse = ", ")`.  The dataset has `r nrow(iris)` observations.  The average petal length is `r mean(iris$Petal.Length)` with a standard deviation (sd) of `r sd(iris$Petal.Length)`.  The average petal width was `r mean(iris$Petal.Width)` (sd = `r sd(iris$Petal.Width)`).  



How to use knitr
========================================================

### Creating the document


- Open RStudio and select `file-> new file -> R markdown`
- Give the document a title (title of the document, not the file) and author
- Specify the output (html, pdf, word).  
    - We will be using html output for this class




Output files
========================================================  

Rstudio with knitr allows your document to be output as:

1. html
2. pdf
3. word document



### Slides using RStudio and knitr

- All slides for this course are made using knitr, which means that all code is dynamically included in the document
- This allows changes in the dataset or code to propagate through to the slides
- It also allows others to look at the underlying code and results


R presentations using knitr
========================================================

We can create R presentations using knitr


### Creating the presentation

- Open RStudio and select `file-> new file -> R presentation`
- Save the file
- Title followed by string of equal signs indicates a new slide
- Can use markdown and introduce code chunks as in an R Markdown document
- Great for teaching





    
- External code
- Output in PDF and Word
- Introduction to functions

More markdown
========================================================


- Can create linked urls: <http://jennakrall.com/IntrotoRepi/>
- Header options
    - theme and highlight
    - table of contents and numbered sections
- More code chunk options
    - global options
    - caching



Markdown/knitr practice
========================================================

Final R exercise

- Goal: Use the diabetes dataset to perform exploratory analyses and apply statistical methods to test scientific hypotheses.
- Option: recreate the document knitr-md-practice.html


