---
title: "Introduction to R: Part 1-B"
author: "Jenna Krall, PhD"
date: "January 2016"
height: 900
font-family: Helvetica
width: 1440
fontsize: 10pt
output:
  beamer_presentation:
    includes:
      in_header: images/header.tex
---


```{r set_knitr_chunk_options, echo = F, message = F, warning = F}
# Source file with setup
source("sourcefile.R")
```



Modules Part 1
========================================================

1. R basics
2. Data frames
3. Plotting
4. Data cleaning basics
5. Statistical tests and models  





Module 1-4
========================================================

Data cleaning basics

1. Reading data into R
2. Dates
3. Matrices
4. Missing data
5. Naming objects



Getting data into R
========================================================
  R can read in data from many different sources

Types of data files

- R data files: (extension .RData or .rda)
- CSV file: comma separated values (extension .csv).
- Excel file: Microsoft Office (extension .xls or .xlsx).  
    - Can use excel to create a .csv file or read into R directly.
- SAS: datafiles (extension .sas7bdat) 

**You must either specify the path to the data or set your working directory to the where the data are located**







  Getting data into R
========================================================

### Google flu trends 

- Estimates flu activity using google searches
- This dataset includes estimated flu activity (by day) for the whole United States, Georgia, Atlanta, and Health and Human Services Region 4 (which includes Alabama, Florida, Georgia, Kentucky, Mississippi, North Carolina, South Carolina, and Tennessee).
- Google's Nature paper: \url{http://www.nature.com/nature/journal/v457/n7232/full/nature07634.html}
- Science paper: \url{http://www.sciencemag.org/content/343/6176/1203}
- Data Source: Google Flu Trends (\url{http://www.google.org/flutrends}), accessed 12/3/2014
- "What we can learn from the epic failure of Google flu trends", (\url{http://www.wired.com/2015/10/can-learn-epic-failure-google-flu-trends/}), accessed 12/18/2015 

Getting data into R
========================================================
  Example using .RData or .rda files
```{r mod4, echo = F}
rm(list = ls())
```
```{r, echo = T}
load("googleflu.RData")
ls()
head(flu)
class(flu)
```



Getting data into R
========================================================
  Example using .csv file
```{r, echo = T}
flu <- read.csv("googleflu.csv", stringsAsFactors = FALSE)
head(flu)
class(flu)
```




Exercise
========================================================

Exercise 4a

- Read in the .csv file "vlbw.csv"




Getting data into R
========================================================
Example using .xls or .xlsx file

- Package `XLConnect` 
- Warning: other packages may not handle dates correctly

```{r, echo = T, eval = F}
wkbook_flu <- loadWorkbook("googleflu.xlsx")
class(wkbook_flu)
flu <- readWorksheet(wkbook_flu, 1)
```


Getting data into R
========================================================
Example using .xls or .xlsx file

```{r}
head(flu)
class(flu)
```



Getting data into R
========================================================
Example using SAS

1. Save SAS data as .csv and read into R
2. Read SAS data directly
    - Only if SAS is installed
    - Do not recommend



Getting data into R
========================================================
1) Save SAS data as .csv and read into R

<br>

- Save data (.sas7bdat) in SAS as .csv file

```
proc export data = googleflu
  outfile = "googleflu.csv"
  dbms = csv replace;
  putnames = yes;
run;
```
- Then use `read.csv` function in R

```{r, eval = F}
flu <- read.csv("googleflu.csv", stringsAsFactors = FALSE)
```








Getting data into R
========================================================
2) Read SAS data directly (only if SAS is installed)

- Need Windows/SAS
- Need R package `sas7bdat` (experimental)
- Use `read.sas7bdat` function to read `.sas7bdat` file directly into R
    
```{r, eval = F}
file1 <- "C:/Users/Desktop/sasfile.sas7bdat"
flu <- read.sas7bdat(file1)
```



Getting data into R
========================================================

Other functions to read in data include

- read.table
- read.delim
- readLines
- scan
- many others





Saving data
========================================================

- Objects in your workspace (console) are not saved
- Saving your workspace (the prompt when closing R)
    - Saves everything in your current workspace (everything shown when call ls())
    - Resulting file is an **.RData file** 
- Saving only relevant output
```{r, eval = F}
save(flu, file = "revised_flu.RData")
```

    - Will overwite if your file name already exists
    - Remember to set your working directory
    
- Save your code as a **.R file** (in the editor)


Saving data
========================================================

Writing as a csv or text file
```{r, eval = F}
write.table(flu, file = "revised_flu.txt", row.names = F)
write.csv(flu, file = "revised_flu.csv", row.names = F)
```




Dates
========================================================

```{r}
load("googleflu.RData")
dates <- flu$Date
head(dates)
class(dates)
dates[2] - dates[1]
```


Dates
========================================================
What if R doesn't know we have a date?
```{r}
flu <- read.csv("googleflu.csv", stringsAsFactors = F)
dates <- flu$Date
head(dates)
class(dates)
```




Dates
========================================================
Use the function `as.Date`

```{r}
dates <- as.Date(dates, format = "%Y-%m-%d")
dates[2] - dates[1]
```


**Until you reassign the date variable within the flu dataset, the flu dataset will not have changed**

```{r}
class(flu$Date)

```


Dates
========================================================

Revise the flu dataset
```{r}
flu$Date <- dates
class(flu$Date)
```

Alternatively, using `mutate`
```{r}
flu <- mutate(flu, Date = as.Date(Date, format = "%Y-%m-%d"))
```




Factors in R
========================================================
    
```{r, echo = F}
vlbw <- read.csv("vlbw.csv", stringsAsFactors = F)
```
    
- Categorical variables
- Take on a finite number of levels
- May be represented as character variables 
```{r}
head(vlbw$sex)
```
or numeric variables 
```{r}
head(airquality$Month)
```

Factors in R
========================================================
    
What are not factor (categorical) variables?

- What did you eat last night?
- Tell us why you are interested in this study?
- What is your reason for visiting the emergency department?


Factors in R
========================================================
    
```{r}
educ <- c("bachelors", "less HS", "some college", "HS", "HS", 
    "some college", "bachelors")
educ
class(educ)
feduc <- factor(educ)
feduc
class(feduc)
```



Factors in R
========================================================
    
Factors in R
```{r}
levels(feduc)
feduc <- factor(feduc, levels = c("less HS", "HS", "some college",
    "bachelors"))
levels(feduc)
feduc <- factor(feduc, levels = c("less HS", "HS", "some college",
    "bachelors"), labels = c("< HS", "HS", "Some college", "BA/BS"))
levels(feduc)
```


Factors in R
========================================================

Iris dataset
```{r}
class(iris$Species)
levels(iris$Species)
```

Matrices and data frames
========================================================

From the R `datasets` package:

- State statistics for the US from the 1970s
```{r}
head(state.x77)
class(state.x77)
```


Matrices and data frames
========================================================


```{r}
str(state.x77)
dim(state.x77)
```


Matrices and data frames
========================================================


```{r}
colnames(state.x77)
rownames(state.x77)
```




Matrices and data frames
========================================================

Cannot select using `$` from matrices
```{r, eval = F}
# Does not work
state.x77$Illiteracy
```






Subsetting vectors in R
========================================================

Recall that we can select items from a vector using brackets

```{r, size = 'tiny'}
x <- c("item 1", "item 2", "item 3", "item 4")
```
We can subset vectors (select elements) using brackets:
```{r}
x[c(1, 3, 4)]
```

We can also use the same techniques to remove items
```{r}
x[-c(1, 4)]
x[-(1 : 3)]
```



Subsetting vectors in R
========================================================
```{r}
x
x == "item 2"
x[x == "item 2"]
newvector <- c(2, 5, 2, 6, 7, 6, 4, 10)
newvector < 5
newvector[newvector < 5]
```


Exercise 
========================================================

Exercise 4b

- Create a vector of 10 random normals using `rnorm`
- Subset the vector to return only positive values




Matrices and data frames
========================================================

Creating a matrix
```{r}
mat1 <- matrix(1 : 12, nrow = 3, ncol = 4)
mat1
```





Matrices and data frames
========================================================


```{r}
mat1[1, 2]
mat1[1 : 2, c(2, 4)]
```





Exercise
========================================================

Exercise 4c

- Create a vector of 20 random normals
- Create a matrix of the vector that is 4 rows and 5 columns
- Extract the third row of the matrix
- Extract the element in the fourth row and second column




Selecting columns from matrices
========================================================

Cannot use "$" to subset matrices
```{r}
head(state.x77[, 1])
head(state.x77[1, ])
head(state.x77[c(2, 3), ])
```




Selecting columns from matrices
========================================================

Cannot use "$" to subset matrices
```{r}
head(state.x77[, "Population"])
head(state.x77["Alabama", ])
head(state.x77["Georgia", "Illiteracy"])
```











Missing data
========================================================

- `NA` is an R symbol used to denote missing values
  - When we take the mean of vectors including missing values, we need to remove those missing values
    ```{r}
vector1 <- c(2, 5, NA, 10, NA, 1, 1, 2.5, 9, 2)
mean(vector1)
mean(vector1, na.rm = TRUE)
```


Missing data
========================================================
We can use `is.na` to determine which elements are missing
```{r}
is.na(vector1)
!is.na(vector1)
vector1[is.na(vector1)]
```



Missing data in data frames and matrices
========================================================


```{r}
mat1
is.na(mat1)
```


Missing data in data frames and matrices
========================================================


```{r}
mat1[2, 3] <- NA
mat1
which(is.na(mat1))
which(is.na(mat1), arr.ind = T)
```



Missing data in data frames and matrices
========================================================


```{r}
complete.cases(mat1)
table(complete.cases(mat1))
mat1[complete.cases(mat1), ]
```




Exercise
========================================================

Exercise 4d

- Read in the dataset `googleflumissing.txt`
- Create a new dataset without any missing data
- Save the new dataset as `googleflucomplete.txt`









Creating new variables
========================================================
"Initializing" a new vector or matrix
```{r}
mat1 <- matrix(nrow = 2, ncol = 2)
mat1
```


Creating new variables
========================================================
Filling in values
```{r}
mat1
mat1[1, 1] <- 1
mat1[1, 2] <- 2
mat1[2, 1] <- 3
mat1[2, 2] <- 4
mat1
```


Creating new variables
========================================================

Creating a dataframe
```{r}
class(mat1)
mat1 <- data.frame(mat1)
class(mat1)
mat1
```



Filtering in base R
========================================================
```{r}
head(flu$Date)
flu$Date[flu$Georgia > 6000]
```

Filtering in base R
========================================================
Recall, in `dplyr`

```{r}
f1 <- filter(flu, Georgia > 6000)
s1 <- select(f1, Date)
head(s1)
head(s1$Date)
```

Filtering in base R
========================================================
Multiple conditions
```{r}
flu$Date[flu$Georgia > 6000 & flu$Atlanta > 6000]
```







Naming objects
========================================================
Suppose we want to add column and rownames to our matrix:
```{r}
mat1 <- matrix(1 : 12, nrow = 3, ncol = 4)
mat1
rownames(mat1)
colnames(mat1)
```



Naming objects
========================================================

We can then assign names:
```{r}
rownames(mat1) <- c("Day1", "Day2", "Day3")
colnames(mat1) <- c("Variable1", "Variable2", "Variable3", 
    "Variable4")
mat1
```

Naming objects
========================================================

Use paste and sequence as a shortcut:
```{r}
rownames(mat1) <- paste0("Day", 1 : 3)
colnames(mat1) <- paste0("Variable", 1 : 4)
mat1
```








Module 1-5
========================================================

Statistical tests and models

1. T-tests
2. Chi-squared tests
3. Relative risk and odds ratio
4. Linear regression and ANOVA
5. Generalized linear models
6. Other models




Data
========================================================

Very low birthweight infant dataset

- "Data on 671 infants with very low (<1600 grams) birth weight from 1981-87 were collected at Duke University Medical Center by Dr. Michael O'Shea, now of Bowman Gray Medical Center."
- O'Shea M, Savitz DA, Hage ML, Feinstein KA: Prenatal events and the risk of subependymal / intraventricular haemorrhage in very low birth weight neonates. Paediatric and Perinatal Epdiemiology 1992;6:352-362.
- Retrieved from \url{http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets}



Data
========================================================

Outcome of interest is intra-ventricular hemorrhage (ivh)

```{r mod5}
vlbw <- read.csv("vlbw.csv", stringsAsFactors = F)
vlbw <- select(vlbw, ivh, bwt, gest, pneumo, delivery, twn)
glimpse(vlbw)
vlbw <- vlbw[complete.cases(vlbw), ]
```



Data
========================================================

Create relevant outcome using dplyr

```{r}
1 * TRUE
head(vlbw$ivh %in% c("definite", "possible"))
head(1 * (vlbw$ivh %in% c("definite", "possible")))
vlbw <- mutate(vlbw, outcome = 1 * ((ivh %in% c("definite", "possible"))))
table(vlbw$ivh, vlbw$outcome)
```



Data
========================================================

*Must be careful if have missing data*

```{r}
NA %in% c("definite", "possible")
```
We will discuss this in more detail later




One sample T tests in R
========================================================

**Review**

- One sample Z and T tests are used for determining whether the mean in a  population is different than a hypothesized value
-  Is the average gestational age for infants born with very low birthweight less than 39 weeks?

Assumptions for Z and T tests

- Large sample size or data are approximately normal if sample size is small

Assumptions for Z test

- Population standard deviation is known



One sample T-tests in R
========================================================
Is average gestational age in the population different than 39 weeks (use $\alpha$=0.05)?

- Null hypothesis $H_0:~\mu=$ 39
- Alternative hypothesis $H_1:~\mu \neq$ 39



One sample T-tests in R
========================================================

```{r}
t_age <- t.test(x = vlbw$gest, mu = 39)
t_age
```


One sample T-tests in R
========================================================

```{r}
class(t_age)
str(t_age)
```



Lists in R
========================================================

```{r}
t_age$statistic
class(t_age$statistic)
t_age[1]
class(t_age[1])
```

Lists in R
========================================================

```{r}
t_age[[1]]
class(t_age[[1]])
t_age[[3]]
t_age$p.value
```






Two sample T-tests
========================================================

**Unpaired two sample t-tests**

Recall that a two sample t-test tests the hypothesis that the means in two populations are the same:

- Does the average gestational age of infants born with very low birthweight differ between males and females?

So we are testing whether the means of a continuous variable differ between two groups:

- Null hypothesis $H_0:~\mu_1= \mu_2$
- Alternative hypothesis $H_1:~\mu_1\neq \mu_2$

If the data are paired, use paired tests





Two sample T-tests
========================================================

Does mean gestational age differ with intra-ventricular hemorrhage?

```{r}
gest_a <- filter(vlbw, delivery == "abdominal")
gest_a <- select(gest_a, gest)
gest_v <- filter(vlbw, delivery == "vaginal") 
gest_v <- select(gest_v, gest)
```

Two sample T-tests
========================================================

Does mean gestational age differ with intra-ventricular hemorrhage?


```{r}
t.test(gest_a, gest_v)
```



Two sample T-tests
========================================================

Does mean gestational age differ with intra-ventricular hemorrhage?


```{r}
t.test(vlbw$gest ~ vlbw$delivery)
```



Exercise
========================================================

Exercise 5a

- Perform a two-sample t-test of whether gestational age is smaller for IVH than no IVH, assuming equal variances

- Null hypothesis $H_0:~\mu_1 \geq \mu_2$
- Alternative hypothesis $H_1:~\mu_1 < \mu_2$





Chi-squared test
========================================================
Are two cateogorical variables independent?

- Is HIV infection associated with MRSA infection?
- Is intra-ventricular hemorrhage associated with delivery type in very low birthweight infants?

Hypothesis:

- Null hypothesis: IVH is independent of delivery type
- Alternative hypothesis: IVH is not independent of delivery type

Assumptions:

- If 2x2 table, no cell counts < 5
- If rxc table, no more than 20% cells < 5






Chi-squared test
========================================================
```{r}
chsq <- chisq.test(vlbw$outcome, vlbw$delivery)
chsq
chsq$p.value
```


Exercise
========================================================

Exercise 5b

- Perform a chi-squared test to see whether the outcome (IVH vs. no IVH) is associated with pneumothorax
- Extract out the test statistic






Relative risk and odds ratio
========================================================
Relative risk (RR)

- Ratio of risks: $p_1$/$p_2$
- Is the risk of disease the same in the exposed and unexposed groups?
- Often interested in testing $H_0:~RR = 1$ vs. $H_1:~RR \neq 1$

Odds ratio (OR)

- Ratio of odds
- Is the odds of disease the same in the exposed and unexposed groups?
- Odds: p/(1-p) or p/q
- OR = ($p_1$/(1 - $p_1$)) / ($p_2$/(1 - $p_2$)) = ($p_1$/$q_1$) / ($p_2$ / $q_2$)
- Often interested in testing $H_0:~OR = 1$ vs. $H_1:~OR \neq 1$






Relative risk and odds ratio
========================================================

- Using R package `epitools`
- Reference groups are first row and first column
```{r}
table_pneumo <- table(vlbw$pneumo, vlbw$outcome)
table_pneumo
```
Relative risk and odds ratio
========================================================

- We want to compare odds of IVH by pneumothorax
```{r}
table_pneumo
epitab(table_pneumo, method = "oddsratio")
```


Relative risk and odds ratio
========================================================


- We want to compare risk IVH by pneumothorax
```{r}
epi_pneumo <- epitab(table_pneumo, method = "riskratio") 
epi_pneumo
```


Relative risk and odds ratio
========================================================
```{r}
str(epi_pneumo)
```













Other R Packages for Epidemiology
========================================================


- Epi
- epiR
- epicalc (no longer maintained)



Sample size calculations in R
======================================================== 

How many observations would we need to test whether a mean is different than some value if

- The difference between the mean and the value is 0.1
- The standard deviation is 1
- We want 90% power

```{r}

power.t.test(delta = 0.1, power = 0.9, type = "one.sample", 
  alternative = "two.sided")
```





Exercise
======================================================== 

Exercise 5d

What is our power to test whether two means are different if

- The difference in means is 2
- The standard deviation is 2
- We want a type I error of 0.1
- We have 10 observations



Sampling data
========================================================

R has the capability of sampling from many distributions:
```{r}
rbinom(10, 1, 0.5)
rnorm(5)
rpois(5, lambda = 5)
rgamma(5, shape = 0.5, rate = 0.5)
```








Simple linear regression
========================================================
Simple linear regression:

$E(y | x) = \beta_0 + \beta_1  x$

where

- $E(y | x)$ is the expectation or mean of $y$
- $\beta_0$ is the intercept, $E(y | x)$ when $x = 0$
- $\beta_1$ is the slope, the change in $E(y | x)$ for a one unit change in $x$

Assumptions

1. Linearity
2. Independence
3. Normality
4. Equal variances



Simple linear regression
========================================================
```{r mod6}
lm1 <- lm(gest ~ delivery, data = vlbw)
lm1
```


Simple linear regression
========================================================
```{r}
str(lm1)
```



Simple linear regression
========================================================

```{r}
summary(lm1)
```



Simple linear regression
========================================================
```{r}
table(vlbw$delivery)
table(vlbw$ivh)
```

Simple linear regression
========================================================

```{r}
lm1 <- lm(gest ~ ivh, data = vlbw)
summary(lm1)

```



Simple linear regression
========================================================

```{r, eval = F}
plot(lm1)
```


Simple linear regression
========================================================

```{r}
vlbw$ivh <- factor(vlbw$ivh, levels = c("definite", 
    "possible", "absent"))
lm1 <- lm(gest ~ ivh, data = vlbw)
summary(lm1)
```



Multiple linear regression
========================================================
Allows us to estimate the association between a predictor and outcome while controlling for other variables

$E(y | x) = \beta_0 + \beta_1  x_1 + \beta_2 x_2 + ... + \beta_P x_P$

where

- $E(y | x)$ is the expectation (mean) of $y$
- $\beta_0$ is the intercept or the $E(y | x)$ when $x_1 = x_2 = ... = x_P = 0$
- $\beta_1$ is the slope, or the change in $E(y | x)$ for a one unit change in $x_1$ when all other covariates ($x_2, ..., x_P$) are held constant.




Multiple linear regression
========================================================

```{r}
lm1 <- lm(gest ~ ivh + bwt, data = vlbw)
sum1 <- summary(lm1)
sum1
```

Multiple linear regression
========================================================



Relevant statistics
```{r}
sum1$r.squared
sum1$adj.r.squared
sum1$fstatistic
```



Exercise
========================================================

Exercise 5d

- Use the output from the model above to find the residuals and create a histogram









Multiple linear regression
========================================================


```{r}
head(lm1$fitted)
newdat1 <- data_frame(c("definite", "possible"), c(1000, 800))
colnames(newdat1) <- c("ivh", "bwt")
newdat1
```



Multiple linear regression
========================================================

```{r}
lm1
predict(lm1, newdata = newdat1)
```



Exercise
========================================================

Exercise 5e

- What is the expected gestational age when IVH is definite and birthweight is 1300?




Multiple linear regression
========================================================

Computing confidence intervals
```{r}
confint(lm1)
```

Interactions
========================================================

Automatically includes main effects
```{r}
lm1 <- lm(gest ~ bwt * delivery, data = vlbw)
summary(lm1)
```


Interactions
========================================================

No main effect terms:
```{r}
lm1 <- lm(gest ~ bwt : delivery, data = vlbw)
summary(lm1)
```

Interactions
========================================================

```{r}
lm1 <- lm(gest ~ ivh * delivery, data = vlbw)
summary(lm1)
```


Removing intercept
========================================================

```{r}
lm1 <- lm(gest ~ -1 + bwt , data = vlbw)
summary(lm1)
```








Analysis of Variance (ANOVA)
========================================================

- ANOVA tests to see whether a continuous variable differs across levels of another variable
- ANOVA tests the "global" null hypothesis for $P$ different groups:
    - Null hypothesis $H_0:~\mu_1= \mu_2 = \mu_3 = ... = \mu_P = \mu$
    - Alternative hypothesis $H_1:$ at least one group not equal

Assumptions:

- Normality
- Independence
- Equal variance: Variance is the same in each of the groups





Analysis of Variance (ANOVA)
========================================================


```{r}
lm1 <- lm(gest ~  ivh, data = vlbw) 
anova(lm1)
```










Generalized linear models
======================================================== 


```{r}
lm(gest ~ ivh, data = vlbw)
glm(gest ~ ivh, data = vlbw)
```






Generalized linear models
======================================================== 


```{r}
glm1 <- glm(gest ~ ivh, data = vlbw, family = "gaussian")
summary(glm1)
```



Exercise
======================================================== 

Exercise 5f

- Fit a logistic regression model to estimate the associations between outcome and delivery (Hint: look at the help page for `family`)










Other models
======================================================== 
R can also be used to fit more complex models including

- Mixed and random effects models
- Survival models
- Ordinal logistic regression
- Time series models
- Bayesian modeling

Helpful resource: \url{http://www.statmethods.net/stats/index.html}










Exercise 
======================================================== 

Diabetes dataset

- Willems JP, Saunders JT, DE Hunt, JB Schorling: Prevalence of coronary heart disease risk factors among rural blacks: A community-based study. Southern Medical Journal 90:814-820; 1997 
- Schorling JB, Roach J, Siegel M, Baturka N, Hunt DE, Guterbock TM, Stewart HL: A trial of church-based smoking cessation interventions for rural African Americans. Preventive Medicine 26:92-101; 1997.
- Retrieved from \url{http://biostat.mc.vanderbilt.edu/wiki/Main/DataSets}



Overview exercise 2
======================================================== 



- Read in the file "diabetes.csv"
- Fit a multiple logistic regression model for the diabetes variable (diab1) on chol, hdl, age, gender, height, weight, waist, and hip.
- Compute the odds ratios corresponding to each variable
- Use `confint` to obtain the lower and upper 95% confidence limits for the odds ratios 





```{r, echo = F}
diab <- read.csv("diabetes.csv")
lm1 <- glm(diab1 ~ chol + hdl + age + gender + height + weight + 
    waist + hip, data = diab, family = "binomial")
# Removing intercept
ci1 <- exp(confint(lm1)[-1, ])
or <- exp(lm1$coef[-1])

```





Plotting confidence intervals
======================================================== 

```{r}
plot(1 : 8, or, pch = 16, 
     ylim = c(0.15, 1.5),
     ylab = "Odds ratio", xlab = "")
segments(x0 = 1 : 8, y0 = ci1[, 1], y1 = ci1[, 2])
abline(h = 1, lty = 2, col = "grey")
```






