---
title: "Introduction to R: Part 1-A"
author: "Jenna Krall, PhD"
date: "January 2016"
height: 900
font-family: Helvetica
width: 1440
fontsize: 10pt
output:
  beamer_presentation:
    includes:
      in_header: images/header.tex
---




```{r set_knitr_chunk_options, echo = F, message = F, warning = F}
# Source file with setup
source("sourcefile.R")
```



Introduction to R
========================================================
Course details

- Course website: \url{http://www.jennakrall.com/cdc}

Instructor details

- Instructor: Dr. Jenna Krall, PhD
- E-mail: jenna.krall@emory.edu


Introduction to R
========================================================


- Course divided into modules
- Lecture materials contain R code
- Exercises throughout modules (~ 5 minutes each)
- Longer overview exercises (~ 15 minutes each)







Introduction to R 
========================================================
Course objectives

- Read in and clean data
- Compute summary statistics
- Apply statistical methods
- Create publication-quality figures
- Create well-formatted output tables
- Create reproducible reports




Modules Part 1
========================================================

1. R basics
2. Data frames
3. Plotting
4. Data cleaning basics
5. Statistical tests and models  





Module 1-1
========================================================

R basics

1. History of R
2. R vs. other statistical programming languages
3. Orientation to R
4. R basics
6. R style 
7. Resources



History of R
========================================================


- Programming language for statistical computing
- Derived from S language (John Chambers, Bell labs)
- GNU (Gnu's not UNIX) project: free and open source
- Developed by
    - Ross Ihaka (University of Auckland)
    - Robert Gentleman (Genentech)
- R can do almost anything
    - Users can create their own R packages
- R project: http://www.r-project.org/
- Comprehensive R Archive Network (CRAN): http://cran.r-project.org/


R vs. other statistical programming languages
========================================================

- The R language
    - We can create objects (e.g. a vector of heights)
    - Objects can interact with other objects (e.g. comparing height and age)
    - Programs consist of applying methods to objects (e.g. what is the mean height?)
- R is interactive
- R and RStudio
- R is easily customizable
    - Customize figures
    - Make nicely formatted tables
    - Write functions




R vs. other statistical programming languages
========================================================
SAS

- Not designed for interactive programming
- Easy to look at your data
- Difficult to create customized output
- Need a lot of code to perform simple tasks (e.g. to find a mean)


In SAS,

```
proc means data = data1;
var variable1;
run;
```
In R,

```{r mod1, eval = F}
mean(variable1)
```




R vs. other statistical programming languages
========================================================

- Stata
    - One dataset at a time
    - Intuitive and easy data cleaning
- MATLAB
    - Matrix computation
    - Powerful
- SPSS
    - Intuitive
    - No programming required



Orientation to R
========================================================

The console

- Type code interactively
- Do not plan to save this code

Editor

- Save code for later
- Can run in console from editor


Orientation to R
========================================================

R vs. RStudio

- R
    - Console and editor
- RStudio
    - Integrated development environment (IDE) for R
    - Runs usual R
    - Nice layout with different panels for editor, console, etc.
    - Additional features




Orientation to R
========================================================


RStudio basics

- Code viewer
- Console
- **History: what commands have been run in current R session**
- **Environment: what is in current workspace** 
- **Files: what files are available on computer**
- Help: view help pages
- Plots
- Packages: what packages are available and loaded
- Viewer: web content




R basics
========================================================

```{r}
5 + 3
```

```{r}
123 / 2 * (2 + 17.2)
```




R basics
========================================================
Can store results using assignment operators

```{r}
nameofobject <- 5
nameofobject
```

Can calculate result as object:
```{r}
nameofsum <- 5 + 3
nameofsum
```
Objects can be functions of other objects:
```{r}
nameofnew <- nameofobject + nameofsum
nameofnew
```




Orientation to R
========================================================


RStudio basics

- Code viewer
- Console
- **History: what commands have been run in current R session**
- **Environment: what is in current workspace** 
- **Files: what files are available on computer**
- Help: view help pages
- Plots
- Packages: what packages are available and loaded
- Viewer: web content


R basics
========================================================
Objects can be reassigned

   - Recall that we created an R object `nameofobject`
```{r}
nameofobject
```
   - Suppose we reassign the name
```{r}
nameofobject <- 1500
nameofobject

```





Functions
========================================================

- Syntax is `functionname(functionarguments)`

```{r}
length(nameofobject)
class(nameofobject)
```




Classes
========================================================

- So far, all the R objects we have dealt with have been numeric  
- There are different data classes that R objects can belong to
- Character objects are non-numeric
- Use quotations to create character objects

```{r}
nameofstring <- "IntrotoR"
nameofstring
class(nameofstring)
```


Classes
========================================================

- Can create vectors by combining numbers or text.  
- The `c` function in R combines objects together.
- **One of the most important functions in your R arsenal**

```{r}
vector1 <- c(6, 2, 3, 4)
vector1
class(vector1)
length(vector1)
```


Classes
========================================================

- Can create integer vectors using a colon

```{r}
2 : 5
```

- Or the `seq` function

```{r}
seq(2, 5)
seq(1, 10, by = 2)
```


Classes
========================================================

Vectors can also consist of strings


```{r, size="footnotesize"}
vectorstring <- c("R", "is", "useful", "for", "data", "analysis")
vectorstring
class(vectorstring)
length(vectorstring)
```

Classes
========================================================
Mixing data classes
```{r}
vectormix <- c("Intro", "to", "R", "version", 1)
vectormix
class(vectormix)
length(vectormix)
```
- Numbers with quotations are not numbers


Classes
========================================================

Why does class matter?

- Functions will perform differently for different classes
    - Mean of a character vector is meaningless

- There are many other classes
    - Later: factors, matrices, data frames, lists, dates
    - Classes for output (e.g. linear model class lm)
    - In R, you can create your own classes of R objects






Selecting elements of a vector
========================================================

Use brackets to select elements of a vector
```{r}
vectorstring
vectorstring[1]
vectorstring[c(1, 4, 5)]
```


Exercise
========================================================

Exercise 1a

- Select only the odd ordered elements of vectorstring (i.e. 1, 3, 5, etc.)







Some useful stat functions
========================================================

```{r}
vector1
mean(vector1)
sd(vector1)
median(vector1)
summary(vector1)
```










Exercise
========================================================
Exercise 1b

- Create a vector of 100 draws from the standard normal distribution
```{r}
vec1 <- rnorm(100)
```
- Compute the summary of the vector
- Create a new vector that contains the first quantile, the median, and the third quantile


Setting the seed
========================================================

- To obtain the *exact* same result for a random process, set the seed beforehand
- R uses a pseudo-random number generator from the starting seed
```{r}
rnorm(10)
set.seed(2371)
rnorm(10)
set.seed(2371)
rnorm(10)
```





Packages 
========================================================

Packages include tools to perform many different analyses

- Some packages are preloaded
    - base (`mean`, `print`, `length`), stats (`median`, `lm`)
    - datasets
    - graphics

- Can install and load additional packages from CRAN
    - ggplot2, RColorBrewer (advanced graphics)
    - dplyr, tidyr (data cleaning/manipulation tools)

Installing a package (may need to specify a CRAN mirror)

```{r, eval = F}
install.packages("ggplot2")
```
Loading a package
```{r, eval = F}
library(ggplot2)
```

Working directory
========================================================

Where R

    - looks for data
    - saves data, figures, etc.

```{r}
getwd()
```



Working directory
========================================================

Change your working directory using `setwd`
```{r, eval = F}
setwd("/home/jkrall/Dropbox/")
```
- For Windows users, setwd("C:/Users/yourusername/etc.")
    - NOTE: must use forward slash!
- For Mac users, setwd("/Users/yourusername/etc.")




Exercise
========================================================

Exercise 1c

- Set your working directory to the folder containing all the class datasets (introR-shortcourse/files)
- Type in the R console `list.files()` to see what is in the current working directory
- Type in the R console `ls()` to see what is in your current R environment



R style 
========================================================

- Appropriate spacing
- Return lines longer than 80 characters (after comma)
```{r, echo = F}
vector1 <- c(6, 2, 3, 4)
```
- Commenting your code
```{r}
# This is the mean of our numeric vector
mean(vector1)
```

> Your closest collaborator is you six months ago but you don’t reply to email.
-- Erin Jonaitis (via Andrew Gelman)




R style 
========================================================


Style guides:

- Hadley Wickham: \url{http://r-pkgs.had.co.nz/style.html}
- Google's: \url{https://google-styleguide.googlecode.com/svn/trunk/Rguide.xml}



Resources
========================================================

R help files (use `?function`)

```
?mean
help.search("mean")
```
Books

- Modern Applied Statistics with S (Venables and Ripley, 2002)
- Advanced R (Wickham, 2014)
- R Programming for Data Science (Peng, 2015) (cheap, available online)
    - <https://leanpub.com/rprogramming>


Resources
========================================================

Search engines

- Google: Append CRAN onto your google search for R.
- R seek: \url{http://www.rseek.org/}

Other resources \url{http://www.jennakrall.com/IntrotoRepi/resources.html}

- R short-refcard: \url{https://cran.r-project.org/doc/contrib/Short-refcard.pdf}
- Stack Overflow: \url{http://stackoverflow.com/questions/tagged/r}
- UCLA R stats: \url{http://www.ats.ucla.edu/stat/r/}
- RStudio cheatsheets: \url{https://www.rstudio.com/resources/cheatsheets/}








Saving 
========================================================

- Objects in your workspace (console) are not saved
- Saving your workspace (the prompt when closing R)
    - Saves everything in your current workspace (everything shown in your Environment tab in RStudio, or alternatively when you call ls())
    - Resulting file is an **.RData file** (more later)
- Save your code as a **.R file** (in the editor)




















Module 1-2
========================================================

Data frames

1. Introduction to `dplyr`
2. Relational operators
3. Dataset manipulation with `dplyr`
4. Data summary statistics


`dplyr`
========================================================

The `dplyr` package for data manipulation

- `dplyr`: A Grammar of Data Manipulation
- Manipulate datasets, including adding and removing columns
- Creating a clean, analytic dataset
- Summarizing your data
- Base R can do this too, but less intuitive



Relational operators
========================================================

- Comparing values
```{r mod2}
2 < 2
2 <= 2
2 == 2
2 != 2
!(2 == 2)
```







Relational operators
========================================================

- Logical vectors
```{r}
x <- rnorm(10)
y <- rnorm(10)
z <- x >= y
x
y
z
class(z)
```



Relational operators
========================================================

- and: `&`
- or: `|`
```{r}
(2 == 2) & (2 == 4)
(2 == 2) | (2 == 4)
```




Relational operators
========================================================

```{r}
2 %in% (1 : 5)
!(2 %in% (1 : 5))
vectorstring
"useful" %in% vectorstring
```


Relational operators
========================================================

- TRUE and FALSE must be always capitalized
- Can use abbreviations (T for TRUE and F for FALSE)

```{r}
F == FALSE
T == TRUE
```


Matrices and data frames
========================================================

- Rows = observations, columns = variables
- Matrices: variables of the same type (e.g. numeric).  Will address later.
- Data frames: mix of numeric, string/character, factor variables

Fisher's Iris dataset

- Data set from Ronald Fisher (1936) "The use of multiple measurements in taxonomic problems"
- Data collected by Edgar Anderson
```{r, eval = F}
?iris
View(iris)
```



Matrices and data frames
========================================================

- **Best command for when you get a new dataset:**

```{r}
str(iris)

```

Fisher's Iris dataset
========================================================

```{r}
head(iris)
class(iris)
```











Subsetting columns from a data frame
========================================================

- If we only want one column (i.e. to take the mean), use `$`

```{r}
seplen <- iris$Sepal.Length
head(seplen)
```



Subsetting columns from a data frame
========================================================

- If we only want one column (i.e. to take the mean), use `$`

```{r}
mean(iris$Sepal.Length)
str(iris$Sepal.Length)
table(iris$Species) 
```






dplyr
========================================================

`dplyr` allows users to clean data frames intuitively (more similar to SAS or Stata)

- Select works to create a new data frame with the selected columns


```{r}
iris_sub <- select(iris, Sepal.Length, Sepal.Width)
head(iris_sub)
```



dplyr
========================================================

- Removing columns

```{r}
iris_sub <- select(iris, -Petal.Length, -Petal.Width, -Species)
head(iris_sub)
```



dplyr vs. SAS
=======================================================
Select works like keep or drop in SAS:
```
data iris; set iris;
  keep Sepal.Length Sepal.Width;
run;

data iris; set iris;
  drop Petal.Length Petal.Width Species;
run;
```


dplyr
========================================================
- Filtering: subsetting and removing rows

```{r}
iris_setosa <- filter(iris, Species == "setosa")
head(iris)
```




dplyr vs. SAS
=======================================================
Filter works like if/where in SAS:

```
data iris; set iris;
  where Species = 'setosa';
run;
```


dplyr
========================================================
Renaming variables

```{r}
head(iris)
iris_sub <- rename(iris, sepallength = Sepal.Length)
head(iris_sub)
```



dplyr vs. SAS
=======================================================
Rename works like rename in SAS (except old and new names are reversed):

```
data iris; set iris;
  rename Sepal.Length = sepallength;
run;

```



dplyr
========================================================

Slice allows you to select specific rows

```{r}
iris_sub <- slice(iris, c(1, 3, 5))
iris_sub
```




Exercise
========================================================

Exercise 2a.

- Create a new dataset of only measurements with petal width less than or equal to 0.5 and only the variables sepal length, sepal width, and petal length






dplyr
========================================================
```{r}
iris
```


dplyr
========================================================

A new type of R object: `tbl_df`

```{r}
iris <- tbl_df(iris)
iris
```



dplyr
========================================================
Arrange is a way of ordering data by many variables easily

```{r}
iris_order <- arrange(iris, Sepal.Length)
iris_order
```



dplyr
========================================================
Adding new variables in `dplyr`
- Much easier than in base R
- Will use frequently in data analysis
```{r}
iris2 <- mutate(iris, slcut = Sepal.Length < 5)
iris2
```

dplyr
========================================================
Adding new variables in `dplyr`
```{r}
iris <- mutate(iris, slcut = Sepal.Length < mean(Sepal.Length))
iris
```




Side note
========================================================

For datasets already in R (via the R `datasets` package), use `data` to revert back to the original dataset

```{r}
data(iris)
head(iris)
```



Exercise
========================================================

Exercise 2b.  

- Create a new variable in the iris dataset that is the standardized sepal length





dplyr
========================================================


`group_by` in `dplyr` specifies grouping variables

```{r}
iris_spec <- group_by(iris, Species)
iris_spec
```






dplyr
========================================================
```{r}
ungroup(iris_spec)
```


dplyr
========================================================

We often want to summarize variables by levels of a factor

- What is the mean Sepal.Length by Species?
- What is the median Sepal.Width by cut sepal length?



dplyr
========================================================

Summarize/summarise allows us to summarize continuous data by factors variables easily
```{r}
summarise(iris_spec, mean1 = mean(Sepal.Length))
```



dplyr
========================================================

Often, we want to summarize multiple variables
```{r}
summarise_each(iris_spec, funs(mean), Sepal.Length, Sepal.Width)
```



dplyr
========================================================

We can also include multiple functions
```{r}
summarise_each(iris_spec, funs(mean, median), Sepal.Length)
```





Exercise
========================================================

Exercise 2c

- Compute the means and standard deviations of all numeric variables in the iris dataset by their Species grouping










dplyr
========================================================
We can also count the number of distinct items in a vector:
```{r}
n_distinct(iris$Species)
table(iris$Species, exclude = NULL)
```




dplyr
========================================================

RStudio cheatsheets: \url{https://www.rstudio.com/resources/cheatsheets/}


\includegraphics[scale=.35]{images/cheat-dplyr.png}










Module 1-3
========================================================

Plotting

1. Introduction to base plotting
2. Customizing plots
3. Saving plots



Introduction to base plotting
========================================================

Base R comes with excellent graphing capabilities

- Scatterplots
- Histograms
- Box plots

Graphical devices (how your computer represents graphical objects) available in R

- pdf
- postscript
- png
- jpeg






Introduction to base plotting
========================================================


- base R plots
    - useful for creating quick plots
    - (easily) customizable
- Other graphics packages exist in R
    - Great for "faceting"
    - Great for multiple panels of plots

Other graphics packages

- ggplot2 
    - Written by Hadley Wickham (RStudio)
    - We will cover ggplot2 later
- lattice
    - Also useful for multiple panels of plots
    - We will not cover in this course







Introduction to base plotting
========================================================

Fisher's `iris` data:
```{r mod3}
head(iris)
head(iris$Sepal.Length)
```




Introduction to base plotting
========================================================
To create a scatterplot, `plot(x, y)`
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width)
```




Introduction to base plotting
========================================================


Histograms show the data distribution for a variable

```{r}
hist(iris$Sepal.Length)
```


Introduction to base plotting
========================================================

- Boxplots show 5-number summary
```{r}
boxplot(iris$Sepal.Length)
```





 
Customizing plots 
======================================================== 
  What can we change? 
 
- Add labels 
- Change colors 
- Change plotting symbol 
- Many other things
```{r, eval = F} 
?plot.default 
?par 
``` 



Short R reference card
======================================================== 

\url{https://cran.r-project.org/doc/contrib/Short-refcard.pdf}

\includegraphics[scale=.7]{images/shortR-plot.pdf}




Customizing plots
========================================================


  Adding labels
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width",
    main = "Scatterplot of Iris data")
```




Colors
========================================================


Use the `col` argument in the `plot` function to set color
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", col = "red",
    main = "Scatterplot of Iris data")
```


Colors
========================================================


We can also use numbers to specify colors based on their order in `palette()`
```{r}
palette()
```



Colors
========================================================


We can also use numbers to specify colors:

```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", col = 3,
    main = "Scatterplot of Iris data")
```



Colors
========================================================


We can also use hexadecimal notation (hex) for the combination of red, green, and blue to specify colors:
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", col = "#FF00FF",
    main = "Scatterplot of Iris data")
```


Colors
========================================================
```{r}
colors()
```




Colors
========================================================

\url{http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf}

<br>
\includegraphics[scale=.4]{images/colors.png}







Colors
========================================================


Plotting by a categorical variable
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", col = iris$Species,
    main = "Scatterplot of Iris data")
```


Plotting symbol
========================================================


```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", pch = 16,
    main = "Scatterplot of Iris data")
```


Plotting symbols
======================================================== 

\includegraphics[scale=.6]{images/plotsymbol.png}



Exercise
========================================================


Exercise 3a

- Plot petal width against sepal width where the color is "dodgerblue" and the plotting symbol is a triangle.





Boxplots
========================================================

- Boxplots show 5-number summary
- Adding boxplot for different values
```{r}
boxplot(iris$Sepal.Length ~ iris$Species)
```



Sizing
========================================================


```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length", 
    ylab = "Sepal width", 
    main = "Scatterplot of Iris data")
```



Sizing
========================================================


Decrease size of plotting points
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", cex = .1,
    main = "Scatterplot of Iris data")
```


Sizing
========================================================


Increase size of axis labels and axes:
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width", cex.lab = 1.5, cex.axis = 2,
    main = "Scatterplot of Iris data")
```







Layering plots
========================================================

Adding a horizontal line at 4:
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     main = "Scatterplot of Iris data")
abline(h = 4, col = "red")
```




Exercise
========================================================

Exercise 3b

- Plot sepal width against sepal length
- Add a horizontal lines for the mean sepal width for each species.


 


Layering plots
========================================================

```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     main = "Scatterplot of Iris data")
abline(h = mean(iris$Sepal.Width), col = "red")
abline(v = mean(iris$Sepal.Length), col = "blue")
```




Layering plots
========================================================

Adding a specific point
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     main = "Scatterplot of Iris data")
points(x = mean(iris$Sepal.Length), y = mean(iris$Sepal.Width), col = "red",
    pch = "+", cex = 2)
```




Layering plots
========================================================

Highlighting specific points
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width,
     main = "Scatterplot of Iris data")
points(x = iris$Sepal.Length[1 : 10], 
    y = iris$Sepal.Width[1 : 10], col = "red")
```






Legends
========================================================

Adding a legend to a plot
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, col = iris$Species,
     main = "Scatterplot of Iris data")
legend("topright", legend = c("setosa", "versicolor", "virginica"), 
       col = 1 : 3, pch = 1)
```


Legends
========================================================

Adding a legend to a plot
```{r}
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, col = iris$Species,
     main = "Scatterplot of Iris data")
legend(5.5, 4, legend = c("setosa", "versicolor", "virginica"), 
      col = 1 : 3, pch = 1)
```





Exercise
========================================================

Exercise 3c

- Modify the previous plot to add horizontal lines at the mean sepal width by species
- Color the lines and points by species
- Add lines to the legend













Saving plots
========================================================

**Set your working directory or specify file path before saving your plots**


Saving your plot as a png (portable network graphic):

- Height and width are in pixels (default is 480 by 480)

```{r, eval = T}
png("Iris_scatterplot.png", height = 700, width = 480)
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width",
    main = "Scatterplot of Iris data")
dev.off()
```

Saving plots
========================================================


Saving your plot as a pdf (portable document format):

- Height and width are in inches (default is 7 by 7)
```{r, eval = T}
pdf("Iris_scatterplot.pdf", height = 11, width = 7)
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width",
    main = "Scatterplot of Iris data")
dev.off()
```




Saving plots
========================================================

What if you don't run `dev.off()`?

- Graphics device does not close
  - You will not have your desired output
    - Can create multiple pages of plots

```{r, results="hide"}
pdf("Iris_scatterplot_2.pdf")
plot(x = iris$Sepal.Length, y = iris$Sepal.Width, xlab = "Sepal length",
    ylab = "Sepal width",
    main = "Scatterplot of Iris data")
plot(x = iris$Petal.Length, y = iris$Petal.Width, xlab = "Petal length",
    ylab = "Petal width",
    main = "Scatterplot of Iris data")
dev.off()
```





Overview exercise 1
========================================================

- Look at the dataset `airquality` in R.
- Create a table of means for each numeric variable (except day and month) by month.  Which variables have missing values?
- Create a histogram of ozone with informative x-axis and main title labels (Hint: look at the help page for the dataset to see the units for ozone).
- Plot ozone against temperature with
    - The color of the plotting points varies by month.
    - Vertical lines corresponding to the mean of temperature for each month.
    - A legend to the top left indicating the color of each month (Hint: the colors R will use by default correspond to the month numbers).

    
The final plot should look like:


Overview exercise 1
========================================================



\includegraphics[scale=.5]{images/overallex1.png}



